---
title: "binostics: computing scagnostics measures in R and C++"
author:
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Hadley Wickham
    affiliation: RStudio, PBC
    email:  hadley@rstudio.com
  - name: Heike Hofmann
    affiliation: Iowa State University
    address: Department of Statistics
    email:  hofmann@iastate.edu
abstract: >
  Scagnostics are useful for exploring large data by helping to find pairs of interesting variables. This paper describes a new implementation available in the binostics package. There are eight scagnostics, outlying, skewed, sparse, clumpy, striated, convex, skinny, stringy, and monotonic. These are computed by processing the results of a convex hull and a minimal spanning tree, in different ways, and the calculations are made efficiently using C++. A data set on Parkinsons disease is used to illustrate the use of the package.
preamble: |
  % Any extra LaTeX you need in the preamble
output: rticles::rjournal_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      root.dir = here::here(), fig.path = here::here("figs/"), 
                      fig.align = 'center', 
                      tidy = TRUE)
library(tidyverse)
library(gridExtra)
```

# Introduction

A scatterplot matrix (SPLOM) is a useful plot to make to examine the distribution of two variables, for linear or nonlinear association, clustering, outliers, discreteness, or barriers. However, if there are many variables there will be too many to practically display. For many variables it can be useful to find the most interesting pairs of variables using scagnostics (scatterplot diagnostics) [@scag,@WW08]. A smaller subset of interesting pairs of variables can then be shown in a SPLOM. 

There are eight measures calculated in scagnostics: outlying, skewed, sparse, clumpy, striated, convex, skinny, stringy, and monotonic. These, themselves can be examined using a SPLOM. This is especially helpful in combination with interactivity, so that the scagnostic values can be explored using mouse-over tooltip, to reveal the pair of variables that generates any value. 

The scagnostics are calculated as a group by the convex hull and computing a minimal spanning tree on the bivariate data. These are processed in different ways to obtain the different metrics. An alternative approach for computing metrics on bivariate data based on correlation are described in @Reshef1518 and available in the \CRANpkg{minerva} package [@minerva]. The archived \CRANpkg{mbgraphic} [@mbgraphic] provides both correlation based and spline-based metrics that are described in [@Grimm2016]. The \CRANpkg{scagnostics} [@LWscagR] provides another implementation that computes scagnostics but it is dependent on the \CRANpkg{rJava} which is not easy to install on all platforms. The \CRANpkg{binostics} package discussed in this paper, implements the calculations using C++. 

There are several alternatives for viewing relationships between more than two variables, including tours [@As85] (available in the \CRANpkg{tourr} package [@tourr]), principal component analysis (several functions available in R), linear discriminant analysis (available in the \CRANpkg{MASS} package [@MASS]), and projection pursuit [@f87]. 

This paper is organized as follows. The next section describes how to compute the scagnostics, and the section following illustrates how they can be used.


# Scagnostics measures

Following @scag the scagnostics measures are evaluated on the data after hexagon binning, and based on the Delaunay triangulation for computational efficiency. In addition, outlying points are removed before computing the measures (with exception of the Outlying measure), to make the measures more robust.

## Underlying definitions
All measures are based on graphs, i.e. a set of vertices and edges, that can all be extracted from the Delaunay triangulation, namely:

- the convex hull, i.e. the outer edges of the Delaunay triangulation
- the alpha hull [@1056714]
- the minimum spanning tree (MST) [@kruskal1956] 

In the following, the length of the MST is defined as the sum of lengths of all edges, and $q_x$ is defined as the $x$th percentile of the MST edge lengths.


When computing the alpha hull, $\alpha$ is set to $q_{90}$ [@scag]. The definition of outlying points is also based on edge length, with points being tagged as outlying if all adjacent edges in the MST have a length larger than
\begin{equation}
w = q_{75} + 1.5 ( q_{75} - q_{25}).
\label{eq:w}
\end{equation}
Several of the measures are corrected for dependence on sample size using the weight
\begin{equation}
\omega = 0.7 + \frac{0.3}{1+t^2},
\end{equation}
where $t = n / 500$ and $n$ the sample size. This correction weight was determined by comparing the scagnostics measures over a large number of datasets [@scag].

## Measure definitions

- **Outlying**: compares the edge lengths of the MST of outlying points with the length of the original MST T 
\begin{equation}
c_{\mathrm{outlying}} = \frac{length(T_{\mathrm{outliers}})}{length(T)}
\end{equation}
- **Skewed**: is measuring skewness in the distribution of MST edge lengths (and thus large values might not correspond to skewed distributions of points) as
\begin{equation}
c'_{\mathrm{skewed}} = \frac{q_{90}-q_{50}}{q_{90}-q_{10}},
\end{equation}
and is corrected to adjust for dependence on the sample size as
\begin{equation}
c_{\mathrm{skewed}} = 1 - \omega (1 - c'_{\mathrm{skewed}}).
\end{equation}
- **Sparse**: detects if points are only found in small number of locations in the plane, as is the case for categorical variables,
\begin{equation}
c_{\mathrm{sparse}} = \omega q_{90}.
\end{equation}
- **Clumpy**: to detect clustering we split the MST in two parts by removing a single edge $j$, and compare the largest edge length within the smaller of the two subsets to the length of the removed edge $j$. The clumpy measure is defined by maximizing over all edges in the MST as
\begin{equation}
c_{\mathrm{clumpy}} = \max_j [1 - \max_k[\mathrm{length}(e_k)] / \mathrm{length}(e_j) ],
\end{equation}
with $k$ running over all edges in the smaller subgraph found after removing edge $j$ from the MST.
- **Striated**: aims to detect patterns like smooth algebraic functions or parallel lines, by evaluating the angles between adjacent edges,
\begin{equation}
c_{\mathrm{striated}} = \frac{1}{|V|} \sum_{v\in V^{(2)}} I(\cos\theta_{e(v,a)e(v,b)} < -0.75),
\end{equation}
where $V$ is the set of vertices, $|V|$ the total number of vertices in the triangulation, $V^{(2)}$ the subset of vertices with two edges (i.e. vertices of degree two), and $I$ the indicator function.
- **Convex**: convexity is computed as the ratio of the area of the alpha hull $A$ and the convex hull $H$, adjusting for sample size dependence,
\begin{equation}
c_{\mathrm{convex}} = \omega \frac{area(A)}{area(H)}.
\end{equation}
- **Skinny**: The ratio of the perimeter to the area of the alpha hull $A$, with normalization such that zero corresponds to a full circle and values close to 1 indicate a skinny polygon,
\begin{equation}
c_{\mathrm{skinny}} = 1 -  \frac{\sqrt{4\pi area(A)}}{perimeter(A)}.
\end{equation}
- **Stringy**: a stringy distribution should have no branches in the MST. This is evaluated by counting the number of vertices of degree two and comparing them to the total number of vertices (dropping those of degree one),
\begin{equation}
c_{\mathrm{stringy}} = \frac{|V^{(2)}|}{|V| - |V^{(1)}|}.
\end{equation}
- **Monotonic**: monotonicity is evaluated via the squared Spearman correlation coefficient, i.e. the Pearson correlation between the ranks of $x$ and $y$,
\begin{equation}
c_{\mathrm{monotonic}} = r_{\mathrm{Spearman}}^2.
\end{equation}



# Interface

The elementary function in the binostics package `scagnostics` and can be called with a pair of vectors or a two-dimensional data structure (a matrix or data frame).

The default S3 method is for a pair of vectors and will compute the scagnostics measures for a single scatter plot. In this case additional control and output is available compared to matrix input. The additional arguments `bins` and `outlierRmv` can be used for detailed checks, but should not be necessary for most applications. The output in this mode is a named list `out`, with `out$s` reporting the scagnostics measures for the scatter plot, and `out$bins` returns the binned data used to compute them. Note that only non-empty bins are kept, and the format is a $3\times n$ matrix where the columns correspond to the bin position in $x, y$ and the bin count, and each row is a non-empty bin. Internally the x and y axis is mapped to integer numbers, and the bin size is reset if there are more than 250 non-empty bins, and this is reflected in the matrix returned to the user.

In most applications we are interested in the scagnostics measures for all combinations of variables in the input data. This is evaluated when passing a matrix or data frame to the `scagnostics` function. In this mode the function returns a data frame, where each row corresponds to one combination of variables, and we have one column for each scagnostics index. Two additional columns `var1` and `var2` report the corresponding variable names.

Finally, we may only be interested in a single scagnostics index for a pair of vectors. Since the measures all rely on the underlying binning and triangulation we simply provide wrapper functions to access this information conveniently. These functions are named after the corresponding scagnostics measure and report the index as an unnamed scalar. This mode would be preferred e.g. when using one of the measures as a projection pursuit index.


# Implementation connecting R and C++

The interface is written in R and handles the reading of the input and finds all combinations of variables in  the case of matrix input. For each pair of variables it then pre-processes the data: entries with missing values in either of the two variables are dropped, and each variable is centered and scaled by its range. It is then using a direct call to C++ for the computation of the measures, as described below, and finally collecting and formatting the output as described above.

The underlying C++ implementation allows for a fast evaluation, in particular it handles the binning and triangulation. The triangulation is used to determine the MST, convex hull and alpha hull, which are then used to compute the measures. The calculation is done in the following steps:

- Binning of the data: we use hexagonal binning, where the number of bins is a free parameter. Note that if this results in more than 250 non-empty bins in the result, half the number of bins will be used instead.
- Computing the Delaunay triangulation and the MST
- Outlier detection: we use a cutoff on the edge length (see Eq. \ref{eq:w}), to identify outlying points. The triangulation and computation of the MST are repeated after removing the tagged outliers.
- Compute scagnostics measures.


# Example

To illustrate the use of scagnostics via the binostics package, we evaluate them for the Parkinsons data [@parkinsons] available from @uci_parkinsons. The dataset is composed of 22 biomedical voice measurements from 31 people, 23 with Parkinson's disease (PD), each column is a particular voice measure, and each row corresponds one of 195 voice recording from these individuals. The original aim of the data was to discriminate healthy people from those with PD, and the data also records the "status" (0 for healthy individuals and 1 for those with PD). Here we instead look for patterns in combinations of the 22 variables, irrespective of the status.

Given the 22 input variables, there are 231 bivariate plots to look at, and we can use scagnostics to select the most interesting ones. We start by reading in the data, dropping the identifier and status columns, and evaluating the scagnostics measures for all variable combinations. To show the data structure we use the `glimpse` function to show the first few entries. The data frame returned by the `binostics::scagnostics` call has one column for each scagnostics measure, and `var1` and `var2` reporting the corresponding variable combination for each entry.

```{r parkinsons}
pk <- as.data.frame(read.csv("data/parkinsons.csv")[,-c(1,18)])
s <- binostics::scagnostics(pk)
tibble::glimpse(s)
```

To get an overview of the results we can make a scatterplot matrix showing pairs of scagnostics measures, where each point now corresponds to one combination of input variables (i.e. one bivariate scatterplot of the original data). This is most useful in an interactive plot that allows us to identify the variables associated with each point, and we can combine the `GGally::ggpairs` [@GGally] with `plotly::ggplotly` [@plotly] to generate this interactive display, as shown in the commented section of the code.

We can generate a scatterplot matrix directly using the data frame returned by `binostics`, where the first nine columns contain the scagnostics measures.

```{r pairs, out.width="100%", fig.width=10, fig.height=10, fig.cap="Scatterplot matrix of scagnostics computed on the parkinsons data."}
# get SPLOM using GGally
GGally::ggpairs(s, columns=1:9)
```

This shows a few outlying points, i.e. parameter combinations that should correspond to interesting scatterplots of the data. To find the corresponding parameters we can use `plotly` in an interactive display. To display the labels when hovering over points we can set the text aesthetics, note however that this will be interpreted as grouping and will result in warnings when generating the graph.


```{r plotly, eval=F}
## read variable columns to generate point labels
## these are only used in the interactive display
s <-  dplyr::mutate(s, labs = paste0(var1,"-",var2))
## to generate a useful interactive plotly version we can
## pass in the labels as text text (note that this will cause warnings in ggpairs)
GGally::ggpairs(s, aes(text = labs), columns=1:9) %>%
  plotly::ggplotly()
```

Some interesting combinations that we have identified are:

- "MDVP.RAP" vs "Jitter.DDP": this combination leads to values close to or exactly one for multiple measures, Striated, Skinny, Stringy and Monotonic, while Convex, Clumpy and Sparse are close to zero. However, the scatterplot is not very interesting, the two variables are perfectly correlated.
- "Shimmer.APQ5" vs "HNR": this combination stand out because it has large Skewed measure (0.84) but lower Monotonic measure (0.53). The scatterplot display shows a branching of observations for larger values of Shimmer.APQ5.
- "MDVP.Fo.Hz." vs "MDVP.Fhi.Hz.": this combination has both low Convex (0.07) and Monotonic (0.25) measures, and the scatterplot shows strong linear correlation with few outlying points.
- "MDVP.Jitter.Abs." vs "Jitter.DDP": this combination stands out as having an unusual combination of measures, including values of Skewed (0.86), Sparse (0.05), Striated (0.18), Skinny (0.41) and Monotonic (0.76). The scatterplot display shows discreetness along MDVP.Jitter.Abs. and an unusual shape of the distribution.

```{r examples, echo=F, out.width="80%", fig.width=8, fig.height=8, fig.cap="Pairs of parkinsons data variables with high or unusual combinations of scagnostic values are shown. "}
p1 <- ggplot(pk, aes(MDVP.RAP, Jitter.DDP)) + geom_point() + theme(aspect.ratio = 1) + ggtitle("A High on Stringy, Monotonic ... ")
p2 <- ggplot(pk, aes(Shimmer.APQ5, HNR)) + geom_point() + theme(aspect.ratio = 1) + ggtitle("B High on Skewed")
p3 <- ggplot(pk, aes(MDVP.Fo.Hz., MDVP.Fhi.Hz.)) + geom_point() + theme(aspect.ratio = 1) + ggtitle("C Low on Convex and Monotonic")
p4 <- ggplot(pk, aes(MDVP.Jitter.Abs., Jitter.DDP)) + geom_point() + theme(aspect.ratio = 1) + ggtitle("D Unusual combination of several")
grid.arrange(p1, p2, p3, p4, nrow=2)
```

# Summary

This paper has described the package \pkg{binostics} which computes scagnostics. Scagnostics are useful when exploring large datasets, in particular for choosing a subset of variables to focus on. The implementation is efficient thanks to the C++ interface, and it is more portable across platforms, unlike the \pkg{scagnostics} package which depends on \pkg{RJava}. 

It would be useful in future work to provide these metrics and the choice of scatterplots to show in an interactive system like a Shiny app [@shiny]. These scagnostics could be usefully combined with projection pursuit to find interesting low-dimensional projections of high-dimensional data, as described in @spinebil. 

# Acknowledgements

The authors gratefully acknowledge the support of the Australian Research Council. The paper was written in \CRANpkg{rmarkdown} [@rmarkdown] using \CRANpkg{knitr} [@knitr].

# Supplementary material 

The source material for this paper is available at \url{https://github.com/uschiLaa/paper-binostics}.

\bibliography{laa-cook-wickham-hofmann}
